---
title: "Réunion - Hydraulic sefety margins"
author: "Anne Baranger"
date: "06/05/2022"
output:
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float: yes
    theme: flatly
  bookdown::word_document2: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "plots/fig_",echo=FALSE)
library(targets)
library(tidyr)
library(dplyr)
library(viridis)
library(terra)
library(ggplot2)
library(sf)
library(data.table)
```

# Introduction

## Pedo-transfer equations

* Campbell: $\Psi_{min} = \Psi_{e} \cdot (SWC_{max}/SWC_{min})^b$
* Van-Genuchten : $\Psi_{min}= [ (\frac{\theta_s - \theta_r}{SWC_{min}-\theta_r})^{1/m}-1]^{1/n} \cdot \frac{1}{\alpha}$
3 computation methods for soil minimum potential.

```{r PTFpars, fig.cap="Pedotransfer function parameters by texture class, subsoils"}
data.frame(texture=c(6,5,4,3,2,1),
            psi_e=c(NA,-0.790569415,-0.9128709202,-1.5811388301,-1.889822365,-5.9761430467),
            b=c(NA,2.6411388301,3.3057418584,4.3822776602,6.5796447301,14.9522860933),
            teta_r=c(0.010,0.0250,0.0100,0.0100,0.0100,0.0100),
            teta_s=c(0.766,0.3660,0.3920,0.4120,0.4810,0.5380),
            alpha=c(0.0130,0.0430,0.0249,0.0082,0.0198,0.0168),
            n=c(1.2039,1.5206,1.1689,1.2179,1.0861,1.0730),
            m=c(0.1694,0.3424,0.1445,0.1789,0.0793,0.0680)) 
```
```{r PTFfun, fig.cap="Campbell and Van Genuchten functions"}
pars_s_VG <- data.frame(texture=c(6,5,4,3,2,1),
            psi_e=c(NA,-0.790569415,-0.9128709202,-1.5811388301,-1.889822365,-5.9761430467),
            b=c(NA,2.6411388301,3.3057418584,4.3822776602,6.5796447301,14.9522860933),
            teta_r=c(0.010,0.0250,0.0100,0.0100,0.0100,0.0100),
            teta_s=c(0.766,0.3660,0.3920,0.4120,0.4810,0.5380),
            alpha=c(0.0130,0.0430,0.0249,0.0082,0.0198,0.0168),
            n=c(1.2039,1.5206,1.1689,1.2179,1.0861,1.0730),
            m=c(0.1694,0.3424,0.1445,0.1789,0.0793,0.0680)) %>% 
    mutate(texture=as.factor(texture))


a=pars_s_VG %>%
  select(texture,psi_e,b,teta_s) %>% 
  crossing(teta=seq(0,1,0.001)) %>% 
  mutate(psi_min=psi_e*((teta_s/teta)^b)) %>% 
  ggplot(aes(teta,psi_min,color=texture))+
  geom_line(size=0.81)+
  ylab("Psi_min (kPa)")+
  xlab("teta")+
  ylim(0,-15000)+
  theme_bw()+
  scale_color_brewer(palette="Spectral",direction=-1)+
  ggtitle("Campbell pedotransfer function")


b=pars_s_VG %>%
  select(texture,teta_r,teta_s,alpha,n,m) %>% 
  crossing(teta=seq(0,1,0.001)) %>% 
  mutate(psi_min=-((((teta_s-teta_r)/(teta-teta_r))^(1/m)-1)^(1/n))*(1/alpha)*9.78*10^(-2)) %>% 
  ggplot(aes(teta,psi_min,color=texture))+
  geom_line(size=0.81)+
  ylab("Psi_min (kPa)")+
  xlab("teta")+
  ylim(0,-15000)+
  theme_bw()+
  geom_vline(xintercept=c(0.01,0.1))+ 
  geom_hline(yintercept = -1000)+
  scale_color_brewer(palette="Spectral",direction=-1)+
  ggtitle("Van-genuchten pdf")

cowplot::plot_grid(a,b,ncol=2)
```

## Soil water content
Extracted from ERA5-land monthly data. Mean of 5 driest month.
Also tested with daily data.

## Texture
Extracted from ESDAC.

# Minimum potential maps

 
```{r SoilPotentialAll,fig.width=10,fig.cap="Minimum soil potential in subsoil for three computation methods: Benito-Garzon, Campbell, Van Genuchten "}
tar_load(psi_min)
##Comparison of Psi_min according to the type of water computation
psi_min %>%
  select(x,y,psi_s_BG,psi_s_CB,psi_s_VG) %>%
    mutate(psi_s_VG=cut(psi_s_VG, breaks=c(-Inf, -50000, -10000, -5000,-3000,-1000,-500,-300,-200,-100,Inf), labels=c("<-50MPa","-50<Psi<-10MPa","-10<Psi<-5MPa","-5<Psi<-3MPa","-3<Psi<-1MPa","-1<Psi<-0.5MPa","-0.5/-0.3","-0.3/-0.2","-0.2/-0.1","-0.1/0")),
         psi_s_CB=cut(psi_s_CB, breaks=c(-Inf, -50000, -10000, -5000,-3000,-1000,-500,-300,-200,-100,Inf), labels=c("<-50MPa","-50<Psi<-10MPa","-10<Psi<-5MPa","-5<Psi<-3MPa","-3<Psi<-1MPa","-1<Psi<-0.5MPa","-0.5/-0.3","-0.3/-0.2","-0.2/-0.1","-0.1/0")),
         psi_s_BG=cut(psi_s_BG, breaks=c(-Inf, -50000, -10000, -5000,-3000,-1000,-500,-300,-200,-100,Inf), labels=c("<-50MPa","-50<Psi<-10MPa","-10<Psi<-5MPa","-5<Psi<-3MPa","-3<Psi<-1MPa","-1<Psi<-0.5MPa","-0.5/-0.3","-0.3/-0.2","-0.2/-0.1","-0.1/0"))) %>% 
  pivot_longer(cols=colnames(.)[c(-1,-2)]) %>%
  ggplot() + # create a plot
  geom_tile( aes(x = x, y = y, fill = value)) + # plot the raw data
  facet_wrap(~name) + # split raster layers up
  theme_bw() +
  scale_fill_brewer(palette="RdYlBu")+
#  scale_fill_discrete(na.value="transparent",type=)+
  labs(x = "Longitude", y = "Latitude",fill="Minimum Potential") + # make plot more readable
  coord_quickmap()
```


# Texture comparison

## Maps
```{r TextureComp}
tar_load(texture)
tar_load(europe)
europe=st_as_sf(europe)

## Texture from ERA5 ##
#######################
texture_ERA=rast("data/texture-2020.nc")
texture_ERA=crop(texture_ERA,vect(europe),mask=TRUE)
texture_ERA=as.data.frame(texture_ERA,xy=TRUE)  %>% 
  filter(slt!=0) %>% 
  mutate(slt=round(slt),
         slt=case_when(slt==1~"coarse",
                      slt==2~"medium",
                      slt==3~"mediumf",
                      slt==4~"fine",
                      slt==5~"vfine",
                      slt==6~"oc",
                      slt==7~"oc"),
         texture=case_when(slt=="coarse"~5,
                           slt=="medium"~4,
                           slt=="mediumf"~3,
                           slt=="fine"~2,
                           slt=="vfine"~1,
                           slt=="oc"~6))
texture_ERA=rast(texture_ERA[,c(1,2,4)],crs="epsg:4326")

## Texture from ESDAC ##
########################
texture_s=texture[,c(1,2,4,6,8)]
colnames(texture_s)=c("x","y","clay","sand","oc")
texture_s=resample(rast(texture_s,crs="epsg:4326"),texture_ERA,method="near")
texture_s <- as.data.frame(texture_s,xy=TRUE) %>% 
  mutate(texture_s=case_when(oc>15~6,
                           clay>60~1,
                           clay>35&clay<=60~2,
                           clay<=35&sand<15~3,
                           clay>18&clay<=35&sand>15~4,
                           clay<=18&sand>15&sand<65~4,
                           clay<=18&sand>=65~5)) %>% 
  mutate(texture_s=as.factor(texture_s))
texture_s=rast(texture_s,crs="epsg:4326")
#texture_ERA=crop(texture_ERA,texture_s)

## Comparison ##
################

comparison=as.data.frame(c(texture_ERA,subset(texture_s,4)),xy=TRUE)
comparison %>% 
  mutate(texture_s=as.factor(texture_s)) %>% 
  ggplot()+
  geom_tile(aes(x=x,y=y,fill=texture_s))+
  geom_sf(data=europe,fill=NA)+
  scale_fill_brewer(palette="Spectral",direction=-1)

comparison %>% 
  mutate(texture=as.factor(texture)) %>% 
  ggplot()+
  geom_tile(aes(x=x,y=y,fill=texture))+
  geom_sf(data=europe,fill=NA)+
  scale_fill_brewer(palette="Spectral",direction=-1)

```

## Potential 

```{r}
## Get data from targets ##
###########################
tar_load(SWC)
SWC=SWC[[1]]

## Compute psi_min ##
#####################
subsoil <- data.frame(texture=c(6,5,4,3,2,1),
                      teta_r=c(0.0100,0.0250,0.0100,0.0100,0.0100,0.0100),
                      teta_s=c(0.766,0.3660,0.3920,0.4120,0.4810,0.5380),
                      alpha=c(0.0130,0.0430,0.0249,0.0082,0.0198,0.0168),
                      n=c(1.2039,1.5206,1.1689,1.2179,1.0861,1.0730),
                      m=c(0.1694,0.3424,0.1445,0.1789,0.0793,0.0680)) 
SWC_text=resample(rast(SWC,crs="epsg:4326"),texture_ERA,method="near")

# psi_min from ERA5 texture #
psi_ERA=as.data.frame(c(texture_ERA,SWC_text),
                      xy=TRUE) %>% 
    left_join(subsoil,by=c("texture")) %>% 
    mutate(psi_ERA_VG=-((((teta_s-teta_r)/(SWC_w_min-teta_r))^(1/m)-1)^(1/n))*(1/alpha)*9.78*10^(-2)) %>% 
    mutate(psi_ERA_VG=case_when(SWC_w_min>teta_s~-0.1,TRUE~psi_ERA_VG))

# psi_min from ESDAC texture # 
psi_text=as.data.frame(c(texture_s,SWC_text),
                      xy=TRUE) %>% 
    left_join(subsoil,by=c("texture_s"="texture")) %>% 
    mutate(psi_s_VG=-((((teta_s-teta_r)/(SWC_w_min-teta_r))^(1/m)-1)^(1/n))*(1/alpha)*9.78*10^(-2)) %>% 
    mutate(psi_s_VG=case_when(SWC_w_min>teta_s~-0.1,TRUE~psi_s_VG))

psi_text=rast(psi_text[,c(1,2,6,14)],crs="epsg:4326")
psi_ERA=rast(psi_ERA[,c(1,2,3,11)],crs="epsg:4326")

as.data.frame(c(psi_ERA,psi_text),xy=TRUE) %>% 
  #filter(psi_ERA_VG<(-15000)) %>% 
  select(x,y,psi_ERA_VG,psi_s_VG) %>% 
  pivot_longer(cols=colnames(.)[c(-1,-2)]) %>% 
  ggplot(aes(x=x,y=y,fill=-value))+
  geom_tile()+
  facet_wrap(~name)+
  scale_fill_gradientn(name = "potential min",trans="log",colours = viridis(80),na.value="transparent")+
  coord_quickmap()

as.data.frame(c(psi_ERA,psi_text),xy=TRUE) %>% 
  select(x,y,psi_ERA_VG,psi_s_VG) %>% 
  filter(psi_ERA_VG>(-10^6)) %>%
  pivot_longer(cols=colnames(.)[c(-1,-2)]) %>% 
  ggplot(aes(-value/1000,color=name))+geom_density()+scale_x_log10()
```


# Soil water content timestep

```{r timestep,fig.cap="Comparison of SWC over four summer months, with daily time step (red) and monthly timestep (blue) measures"}
tar_load(SWC)
SWC_t=SWC[[2]]
france_ext=ext(-5,10,41,52)
SWC_t= crop(rast(SWC_t,crs="epsg:4326"),france_ext)

## Points to extract
toextract <- data.frame(Luberon=c(5.387792,43.814892),
                         Landes= c(-0.438943,44.427567), #Landes
                         Orléans= c(1.9,47.4), #la maison
                         Chamrousse= c(5.860739,45.092439), #Chamrousse
                         GrandEst=c(5.569744,48.475034),
                        Grignols=c(0.0,44.4)) #Grand-Est

loc= colnames(toextract)
row.names(toextract)=c("Lon","Lat")
toextract=t(toextract)
row.names(toextract) <- NULL


## Extract from monthly data
SWC_m=setDT(as.data.frame(SWC_t,xy=TRUE))
SWC_m=SWC_m[,loc_ID:=.I]
SWC_m=melt.data.table(SWC_m,id.vars=c("loc_ID","x","y"))
SWC_m[,time:=as.numeric(sub(".*_","",variable))]
SWC_m[,year:=as.factor(1950+(as.numeric(time)-1)%/%12)]
SWC_m[,month:=as.factor(as.numeric(time)%%12)]
SWC_m=SWC_m[year%in%c(2003,2009,2015,2021),]
SWC_m=SWC_m[month%in%c(6,7,8,9),]
SWC_m=dcast(data=SWC_m,loc_ID+x+y~variable,value.var = c("value"))
SWC_m=rast(SWC_m[,2:19],crs="epsg:4326")

SWC_mtime=as.data.frame(extract(SWC_m,toextract),
                           row.names = loc) %>% 
    tibble::rownames_to_column(var="Loc") %>% 
    pivot_longer(cols=colnames(.)[c(-1)]) %>% 
    separate(name,sep="_",c("drop","time")) %>% 
    select(-drop) %>% 
    mutate(time=as.numeric(time)) 

## Extract from daily data
SWC_d=rast("data/swc-daily.nc")
SWC_d= crop(SWC_d,france_ext,snap="out")
SWC_d <- (7*SWC_d[[grepl("swvl1", names(SWC_d))]]+
  21*SWC_d[[grepl("swvl2", names(SWC_d))]]+
  72*SWC_d[[grepl("swvl3", names(SWC_d))]]+
  189*SWC_d[[grepl("swvl4", names(SWC_d))]])/289
SWC_dtime=as.data.frame(extract(SWC_d,toextract),
                           row.names = loc) %>% 
    tibble::rownames_to_column(var="Loc") %>% 
    pivot_longer(cols=colnames(.)[c(-1)]) %>% 
    separate(name,sep="_",c("drop","time")) %>% 
    select(-drop) %>% 
    mutate(time=as.numeric(time)) 

SWC_mtime=cbind(SWC_mtime,timy=c(15,45,76,0),year=c(rep(1,4),rep(2,4),rep(3,4),rep(4,4)))
cbind(SWC_dtime,year=c(rep(1,92),rep(2,92),rep(3,92),rep(4,92))) %>% 
  mutate(timy=time%%92) %>% 
  left_join(SWC_mtime %>% select(-time),by=c("Loc","year","timy")) %>% 
  select(Loc,time,value.x,value.y) %>% 
  rename("Daily SWC"=value.x,
         "Monthly SWC"=value.y) %>% 
  pivot_longer(cols=c("Daily SWC","Monthly SWC")) %>% 
  ggplot(aes(time,value,color=name))+
  geom_point(size=0.7)+
  facet_wrap(~Loc)+
  theme(legend.position = "bottom")
```
```{r}
cbind(SWC_dtime,year=c(rep(1,92),rep(2,92),rep(3,92),rep(4,92))) %>% 
  filter(Loc=="Grignols") %>% 
  mutate(psi_3=-((((0.366-0.025)/(value-0.025))^(1/0.03424)-1)^(1/n))*(1/alpha)*9.78*10^(-2),
         psi_5=-((((teta_s-teta_r)/(SWC_w_min-teta_r))^(1/m)-1)^(1/n))*(1/alpha)*9.78*10^(-2))
```

